Estimating survival in left filtered data

### 2021-12-11 reports/04-gm.Rmd
Interim results; running on GM data now

@@

In the simulation study, we showed that under several scenarios compatible with our hypothetical causal structure, the AWKM survival estimator had small bias compared to the WKM estimator. The bias was smaller when the cumulative incidence of the outcome was low and at later follow-up time points. Now, we estimate cancer-free survival in a real-world context. Using data from the UAW-GM Cohort study, we follow individuals starting from hire to 40 years after hire for incidence of colon cancer. This duration of follow-up spans the vast majority of individuals' working lifetimes.

Colon cancer incidence was obtained via linkage to the Michigan Cancer Registry, which recorded cancer incidence cases starting on January 1, 1985.
<!-- and linkage to Surveillance, Epidemiology, and End Results (SEER), which recorded cancer incidence cases starting on January 1, 1973. -->
The cohort is comprised of individuals hired between 1938 and 1985. Cancer-free survival to the start of the registry is a left-filtering process possibly in the presence of the HWSE as investigated in the simulations. Over the 40 year follow-up period, vital status was obtained through the Social Security Administration, the National Death Index, as well as records provided by the UAW. Baseline confounders included race, sex, plant, and year of hire. Time-varying confounders included age, cumulative time off, and exposure to the metalworking fluids not under study. Representations of these terms were included in the estimation of both the treatment and censoring mechanisms, which were estimated with stratification on every year of follow-up. The exposure rules of interest were identical to those in the simulation study: $a_0$ never exposed and $a_1$ always exposed while at work. We assume that the effect of exposure on health status, mortality, cancer incidence, etc occurs after a lag of 15 years. Similarly, we assume that the effects of cumulative time off occur after a 15 year lag. Counterfactual survival under rules $a_0$ and $a_1$ were estimated using the WKM and AWKM estimators.

```{r, eval=F}
mwf <- NULL
source(here("scripts", "04-application.R"))
```

>>>>

```{r, include=F}
knitr::opts_chunk$set(echo = F, results = "asis", fig.align = "center")
if (!"mytheme" %in% ls()) {
  source("~/headRs/00-my-theme.R")
  library(data.table)
  options(tikzDefaultEngine = 'pdftex')
}
library(tidyverse); library(data.table)
library(here)
library(tableone)
```

@@

In the simulation study, we showed that under several scenarios compatible with our hypothetical causal structure, the AWKM survival estimator had small bias compared to the WKM estimator. The bias was smaller when the cumulative incidence of the outcome was low and at later follow-up time points. Next, we estimated cancer-free survival in a real-world context. Using data from the UAW-GM Cohort study, we followed $38\,553$ individuals starting from hire to 40 years after hire for incidence of colon cancer. This duration of follow-up spans the vast majority of individuals' working lifetimes. As in the simulation, the UAW-GM data were longitudinal data with baseline covariates, time-varying covariates, and a survival outcome.

The exposures of interest were MWF of three types: straight, soluble, and synthetic [@Byers_2006; @Mirer_2003; @Mirer_2010]. Straight MWFs are hydrocarbon-based fluids whose use became widespread by the 1920s. They continue to be widely used because of their simple formulation. In straight MWFs, hydrocarbons of different lengths are mixed together with other additives to attain different properties. Straight MWFs contain polycyclic aromatic hydrocarbons, long known to be carcinogenic [@IARC_1973]. Soluble oils are water-based oil emulsions first introduced in response to rising oil prices. They now make up the largest market share of MWFs [@Childers_2006]. Soluble oils are vulnerable to microbial contamination, so the addition of biocides is needed. However, their high lubricity makes soluble MWFs the most popular fluid type. Synthetic MWFs have the best toxicological profile, have no oil, and have a higher resistance to microbial growth. They were introduced into the MWF market in the second half of the 20th C., but fail to out-perform soluble MWFs in metalworking applications. Soluble and synthetic MWFs contain biocides, corrosion inhibitors, and chlorinated compounds, some classified as carcinogenic by the IARC [@IARC_1987].

The outcome of interest was colon cancer incidence. There is little past research linking colon cancer to MWF exposures, but there is some evidence suggesting that straight MWFs cause colon cancer [@Izano_2019]. We obtained cancer incidence data via linkage to the Michigan Cancer Registry, which recorded cancer incidence cases starting on January 1, 1985.
<!-- and linkage to Surveillance, Epidemiology, and End Results (SEER), which recorded cancer incidence cases starting on January 1, 1973. -->
The cohort is comprised of individuals hired between 1938 and 1985. Cancer-free survival to the start of the registry was a left-filtering process possibly in the presence of the HWSE as investigated in the simulations. Over the 40 year follow-up period, vital status was obtained through the Social Security Administration, the National Death Index, as well as records provided by the UAW. Baseline confounders included race, sex, plant, and year of hire. Time-varying confounders included age, cumulative time off, and exposure to the metalworking fluids not under study. These terms were included in the estimation of both the treatment and censoring mechanisms, which were estimated with stratification on every year of follow-up. The exposure rules of interest were identical to those in the simulation study: $a_0$ never exposed and $a_1$ always exposed while at work. We assumed that the effect of exposure on health status, mortality, cancer incidence, etc occurs after a lag of 15 years. Similarly, we assumed that the effects of cumulative time off occur after a 15 year lag. Counterfactual survival under rules $a_0$ and $a_1$ were estimated using the WKM and AWKM estimators. Summary statistics for the full study population and the colon cancer cases are presented in Table \ref{tab:popchar}.

```{r, eval=F}
mwf <- NULL
if (!"cohort_analytic" %in% ls()) {source(here("scripts", "04-application.R"))}
full.tab <- cohort_analytic[,.(
	studyno,
	Race,
	Sex,
	Plant,
	yin = yin.gm,
	age = age.year2/365,
	off = off.gan + off.san + off.han,
	straight = as.integer(straight > 0),
	soluble = as.integer(straight > 0),
	synthetic = as.integer(straight > 0),
	death = `All causes`,
	canc_co
)]

nrow(full.tab)
print(
	CreateTableOne(data = full.tab[,.(
		Race = Race[1],
		Sex = Sex[1],
		Plant = c("Plant 1", "Plant 2", "Plant 3")[which.max(table(Plant))],
		Dead = factor(max(death))
	), studyno]),
	showAllLevels = T)

print(
	CreateTableOne(data = full.tab[,.(
		Straight = factor(as.integer(sum(shift(straight, 15, 0)) <= 0)),
		Soluble = factor(as.integer(sum(shift(soluble, 15, 0)) <= 0)),
		Synthetic = factor(as.integer(sum(shift(synthetic, 15, 0)) <= 0)),
		`Year of hire` = yin[1],
		Age = age[min(.N, .I[canc_co == 1], na.rm = T)],
		`Cumulative time off` = shift(
			off, 15, 0)[min(.N, .I[canc_co == 1], na.rm = T)]
	), studyno])
	)

nrow(full.tab[studyno %in% full.tab[canc_co == 1, studyno]])
print(
	CreateTableOne(data = full.tab[studyno %in% full.tab[canc_co == 1, studyno],.(
		Race = Race[1],
		Sex = Sex[1],
		Plant = c("Plant 1", "Plant 2", "Plant 3")[which.max(table(Plant))]
	), studyno]),
	showAllLevels = T)

print(
	CreateTableOne(data = full.tab[studyno %in% full.tab[canc_co == 1, studyno],.(
		Straight = factor(as.integer(sum(shift(straight, 15, 0)) <= 0)),
		Soluble = factor(as.integer(sum(shift(soluble, 15, 0)) <= 0)),
		Synthetic = factor(as.integer(sum(shift(synthetic, 15, 0)) <= 0)),
		`Year of hire` = yin[1],
		Age = age[min(.N, .I[canc_co == 1], na.rm = T)],
		`Cumulative time off` = shift(
			off, 15, 0)[min(.N, .I[canc_co == 1], na.rm = T)]
	), studyno])
	)

xtable(head(cars))
```

\begin{table}[h]
\caption{Study population characteristics.}
\label{tab:popchar}
\begin{center}\begin{tabular}{lrlrl}
\toprule
&		\multicolumn{2}{c}{Full cohort} 	  & \multicolumn{2}{c}{Colon cancer cases}		\\
\midrule
$n$ (person-years) 
            &     38 553 & (1 402 372)       & 267 & (9 635)		\\
Race (\%) \\
\hspace{1.5em} Black   &      7 133 & (18.5)    			&  83 & (31.1)  		\\
\hspace{1.5em} White   &     31 420 & (81.5)    			& 184 & (68.9)  		\\
Sex (\%) \\
\hspace{1.5em} Female  &      4 757 & (12.3)    			&  49 & (18.4)		\\
\hspace{1.5em} Male    &     33 796 & (87.7)    			& 218 & (81.6)		\\
Plant (\%) \\
\hspace{1.5em} Plant 1 &      9 092 & (23.6)    			&  66 & (24.7)		\\
\hspace{1.5em} Plant 2 &     17 090 & (44.3)    			& 107 & (40.1) 		\\
\hspace{1.5em} Plant 3 &     12 371 & (32.1)    			&  94 & (35.2)    		\\
\midrule
Ever exposed to MWF (\%)
						&     22193 & (57.6)    			& 104 & (39.0)  		\\
Year of hire (mean (\textsc{sd}))
						&     1963 & (11.62)    		& 1968 & (8.79) 		\\
Age (mean (\textsc{sd}))
						&     63.38 & (9.08)    			& 67.78 & (9.27) 		\\
Cumulative time off (mean (\textsc{sd}))
						&     0.04 & (0.14)    			  & 0.05 & (0.15) 		\\
\bottomrule
\end{tabular}\end{center}
\end{table}

## Assumptions

Since we are working with observational data, the evaluation of the no-interference, causal consistency, ignorability, and positivity assumptions are critical for causal inference. The stability of our estimation depends on positivity, which we assessed qualitatively by examining the distribution of the weights. The no-interference assumption may be problematized by the fact that there were a finite number of job types in the factory setting. If one worker operates a particular metalworking machine, then the the other workers are not able to operate that machine at that time. Instead, they may be assigned to assembly tasks, which have lower MWF exposure opportunities. That said, since these factories were quite large, there may be approximate independence. The consistency assumption is also problematic. The MWFs of interest are complex chemical mixtures whose composition changes by design and by nature. Over the last several decades, the formulation of MWFs has changed significantly in reaction to performance needs and toxicity concerns [@Mirer_2003; @Byers_2006]. The composition of MWFs also undergoes unintentional changes because of the nature of their use: MWFs are often applied in contexts where contamination by other substances and microbes is possible and chemical changes due to heat and pressure are likely. Indeed, there are substantial concerns over the carcinogenicity of the chemical species formed in the MWF mixtures that were not originally added [@Hidajat_2020]. Concerns regarding the consistency assumption may be abrogated in part by adequate adjustment for secular and factory-level characteristics.

Another key assumption meriting discussion is that of conditional ignorability. In order to achieve identification, even in the absence of left filtering, we need to have conditionally ignorable future exposure status and ignorable future censoring status at each time point given past data. In occupational cohorts, employment status and health history are strong predictors of future death [@Hafner_1987;@Halliday_2014;@Laliotis_2018]. Logically, major causes of death first act through employment status before they precipitate death. This dynamic is actually a key component in the setup for HWSE. We are therefore relatively confident that conditional ignorability of censoring due to death is attained given covariate, exposure, and cancer history. Our confidence in the conditional ignorability of exposure given history was not as strong. In particular, workers may be assigned to certain tasks based on their specific skills and knowledge, which may be associated with structural privileges that confer a lower risk of deleterious health outcomes. The potential magnitude of this uncontrolled confounding is likely to be limited, however. The education level of the workers in the cohort was homogeneous, and all cohort members were members of the UAW union, which had uniform procedures in place for equitable access to training, wages, and career advancement [@Harbison_1950;@Barnard_2005]. The presence of UAW policies support the assertion that given time since hire, job types (and therefore exposures) were randomly allocated.

## Results

- Foo
